---
title: "Exercises for class 1"
output: html_document
date: "13-02-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Answers: https://sr2-solutions.wjakethompson.com/bayesian-inference

#git push origin main
#git remote add upstream https://github.com/methods-4-f24/methods-4-classes
#git fetch upstream
#git merge upstream/main
#git add . 
#git commit -m "comment"
#git push origin main

```

```{r}
install.packages(c("coda","mvtnorm","devtools", "usethis"))
devtools::install_github("rmcelreath/rethinking",ref="Experimental")
```

```{r}
pacman::p_load(dagitty)
library(rethinking)
```

# Exercises for Class 1

## Supporting code

Below is code from Chapter 2 of the book that you can use to do the exercises. Not all the code you require is here though. Make sure to not rush it and take time to play around with each function. :))

```{r}
# Grid Approximation

# define grid
p_grid = seq(from = 0, to = 1, length.out=20)

# define prior
prior = rep(1, 20)
#prior <- ifelse( p_grid < 0.5 , 0 , 1 )
#prior <- exp( -5*abs( p_grid - 0.5 ) )

# compute likelihood at each value in grid
likelihood = dbinom(6, size = 9, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior = likelihood * prior

# standardize the posterior so it sums to 1
posterior = unstd.posterior / sum(unstd.posterior)

# plot the distribution
plot(p_grid, posterior, type = "b",
     xlab = "probability of water", ylab = "posterior_probability")
mtext("20 points")

```

```{r}
## R code 2.6
library(rethinking)
globe.qa <- quap(
    alist(
        W ~ dbinom( W+L ,p) ,  # binomial likelihood
        p ~ dunif(0,1)     # uniform prior
    ) ,
    data=list(W=6,L=3) )

# display summary of quadratic approximation
precis( globe.qa )

## R code 2.7
# analytical calculation
W <- 6
L <- 3
curve( dbeta( x , W+1 , L+1 ) , from=0 , to=1 )
# quadratic approximation
curve( dnorm( x , 0.67 , 0.16 ) , lty=2 , add=TRUE )
```
The function precis presents a brief summary of the quadratic approximation. In this case,
it shows the posterior mean value of p = 0.67, which it calls the “Mean.” The curvature is
labeled “StdDev” This stands for standard deviation. This value is the standard deviation of
the posterior distribution, while the mean value is its peak. Finally, the last two values in the
precis output show the 89% percentile interval, which you’ll learn more about in the next
chapter. You can read this kind of approximation like: Assuming the posterior is Gaussian, it
is maximized at 0.67, and its standard deviation is 0.16.

The blue (full line) curve is the analytical posterior and the black (dotted) curve is the quadratic approximation. The black curve
does alright on its left side, but looks pretty bad on its right side. It even assigns positive

```{r}
## R code 2.8
n_samples <- 1000
p <- rep( NA , n_samples )
p[1] <- 0.5
W <- 6
L <- 3
for ( i in 2:n_samples ) {
    p_new <- rnorm( 1 , p[i-1] , 0.1 )
    if ( p_new < 0 ) p_new <- abs( p_new )
    if ( p_new > 1 ) p_new <- 2 - p_new
    q0 <- dbinom( W , W+L , p[i-1] )
    q1 <- dbinom( W , W+L , p_new )
    p[i] <- ifelse( runif(1) < q1/q0 , p_new , p[i-1] )
}

## R code 2.9
dens( p , xlim=c(0,1) )
curve( dbeta( x , W+1 , L+1 ) , lty=2 , add=TRUE )

```



## Exercises

### Easy

2E1. Which of the expressions below correspond to the statement: the probability of rain on Monday?
(1) Pr(rain)
(2) Pr(rain|Monday)
(3) Pr(Monday|rain)
(4) Pr(rain, Monday) / Pr(Monday)

```{r}
#2 og #4 - kan også skrives sådan:  Pr(Monday|rain) Pr(rain) / Pr(Monday)
```

2E2. Which of the following statements corresponds to the expression: Pr(Monday|rain)?
(1) The probability of rain on Monday.
(2) The probability of rain, given that it is Monday.
(3) The probability that it is Monday, given that it is raining.
(4) The probability that it is Monday and that it is raining.

```{r}
#3
```

2E3. Which of the following expressions below correspond to the statement: the probability that it is Monday, given that it is raining?
(1) Pr(Monday|rain)
(2) Pr(rain|Monday)
(3) Pr(rain|Monday) Pr(Monday)
(4) Pr(rain|Monday) Pr(Monday) / Pr(rain)
(5) Pr(Monday|rain) Pr(rain) / Pr(Monday)

```{r}
#1 og #4
```

2E4. The Bayesian statistician Bruno de Finetti (1906–1985) began his 1973 book on probability theory with the dedication: “PROBABILITY DOES NOT EXIST.” The capitals appeared in the original, so I imagine de Finetti wanted us to shout this statement. What he meant is that probability is a device for describing uncertainty from the perspective of an observer with limited knowledge; it has no objective reality. Discuss the globe tossing example from the chapter, in light of this statement. What does it mean to say “the probability of water is 0.7”?

```{r}

#With the current knowledge, the probability is 0.7, however, this does not apply in the real world, as we for sure do not have all possible evidence.

```

### Medium

2M1. Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p. (1) W, W, W (2) W, W, W, L (3) L, W, W, L, W, W, W

```{r}
# Grid Approximation nr. 1
# define grid
p_grid = seq(from = 0, to = 1, length.out=20)

# define prior
prior = rep(1, 20)

# compute likelihood at each value in grid
likelihood = dbinom(3, size = 3, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior = likelihood * prior

# standardize the posterior so it sums to 1
posterior = unstd.posterior / sum(unstd.posterior)

# plot the distribution
plot(p_grid, posterior, type = "b",
     xlab = "probability of water", ylab = "posterior_probability")
mtext("20 points")


# Grid Approximation nr. 2
# define grid
p_grid = seq(from = 0, to = 1, length.out=20)

# define prior
prior = rep(1, 20)

# compute likelihood at each value in grid
likelihood = dbinom(3, size = 4, prob = p_grid) #3 succes out of 4

# compute product of likelihood and prior
unstd.posterior = likelihood * prior

# standardize the posterior so it sums to 1
posterior = unstd.posterior / sum(unstd.posterior)

# plot the distribution
plot(p_grid, posterior, type = "b",
     xlab = "probability of water", ylab = "posterior_probability")
mtext("20 points")


# Grid Approximation nr. 3
# define grid
p_grid = seq(from = 0, to = 1, length.out=20)

# define prior
prior = rep(1, 20)

# compute likelihood at each value in grid 
likelihood = dbinom(5, size = 7, prob = p_grid) #5 succes of 7

# compute product of likelihood and prior
unstd.posterior = likelihood * prior

# standardize the posterior so it sums to 1
posterior = unstd.posterior / sum(unstd.posterior)

# plot the distribution
plot(p_grid, posterior, type = "b",
     xlab = "probability of water", ylab = "posterior_probability")
mtext("20 points")
```
And here, the tidyverse way:
```{r}
library(tidyverse)


dist <- tibble(p_grid = seq(from = 0, to = 1, length.out = 20),
               prior = rep(1, times = 20)) %>%
  mutate(likelihood_1 = dbinom(3, size = 3, prob = p_grid),
         likelihood_2 = dbinom(3, size = 4, prob = p_grid),
         likelihood_3 = dbinom(5, size = 7, prob = p_grid),
         across(starts_with("likelihood"), ~ .x * prior),
         across(starts_with("likelihood"), ~ .x / sum(.x))) %>%
  pivot_longer(cols = starts_with("likelihood"), names_to = "pattern",
               values_to = "posterior") %>%
  separate(pattern, c(NA, "pattern"), sep = "_", convert = TRUE) %>%
  mutate(obs = case_when(pattern == 1L ~ "W, W, W",
                         pattern == 2L ~ "W, W, W, L",
                         pattern == 3L ~ "L, W, W, L, W, W, W"))

ggplot(dist, aes(x = p_grid, y = posterior)) +
  facet_wrap(vars(fct_inorder(obs)), nrow = 1) +
  geom_line() +
  geom_point() +
  labs(x = "Proportion Water (p)", y = "Posterior Density")
```

2M2. Now assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when p ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of observations in the problem just above.

```{r}
# In the book, McElreath says that we can do much better than choosing a uniform prior for estimating the coverage of water. Is this what we are doing here? How does that impact our estimation?

# define grid
p_grid = seq(from = 0, to = 1, length.out=20)

prob = p_grid

# compute likelihood at each value in grid
likelihood = dbinom(3, size = 3, prob = p_grid)

# define prior
prior = ifelse(prob<0.5,0,1)

# compute product of likelihood and prior
unstd.posterior = likelihood * prior

# standardize the posterior so it sums to 1
posterior = unstd.posterior / sum(unstd.posterior)

# plot the distribution
plot(p_grid, posterior, type = "b",
     xlab = "probability of water", ylab = "posterior_probability")
mtext("20 points")

# define grid
p_grid = seq(from = 0, to = 1, length.out=20)

prob = p_grid

# compute likelihood at each value in grid
likelihood = dbinom(3, size = 4, prob = p_grid)

# define prior
prior = ifelse(prob<0.5,0,1)

# compute product of likelihood and prior
unstd.posterior = likelihood * prior

# standardize the posterior so it sums to 1
posterior = unstd.posterior / sum(unstd.posterior)

# plot the distribution
plot(p_grid, posterior, type = "b",
     xlab = "probability of water", ylab = "posterior_probability")
mtext("20 points")


# define grid
p_grid = seq(from = 0, to = 1, length.out=20)

prob = p_grid

# compute likelihood at each value in grid
likelihood = dbinom(5, size = 7, prob = p_grid)

# define prior
prior = ifelse(prob<0.5,0,1)

# compute product of likelihood and prior
unstd.posterior = likelihood * prior

# standardize the posterior so it sums to 1
posterior = unstd.posterior / sum(unstd.posterior)

# plot the distribution
plot(p_grid, posterior, type = "b",
     xlab = "probability of water", ylab = "posterior_probability")
mtext("20 points")


#The tidyverse one
dist <- tibble(p_grid = seq(from = 0, to = 1, length.out = 20)) %>%
  mutate(prior = case_when(p_grid < 0.5 ~ 0L,
                           TRUE ~ 1L),
         likelihood_1 = dbinom(3, size = 3, prob = p_grid),
         likelihood_2 = dbinom(3, size = 4, prob = p_grid),
         likelihood_3 = dbinom(5, size = 7, prob = p_grid),
         across(starts_with("likelihood"), ~ .x * prior),
         across(starts_with("likelihood"), ~ .x / sum(.x))) %>%
  pivot_longer(cols = starts_with("likelihood"), names_to = "pattern",
               values_to = "posterior") %>%
  separate(pattern, c(NA, "pattern"), sep = "_", convert = TRUE) %>%
  mutate(obs = case_when(pattern == 1L ~ "W, W, W",
                         pattern == 2L ~ "W, W, W, L",
                         pattern == 3L ~ "L, W, W, L, W, W, W"))

ggplot(dist, aes(x = p_grid, y = posterior)) +
  facet_wrap(vars(fct_inorder(obs)), nrow = 1) +
  geom_line() +
  geom_point() +
  labs(x = "Proportion Water (p)", y = "Posterior Density")

```

2M3. Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 70% covered in water. The Mars globe is 100% land. Further suppose that one of these globes—you don’t know which—was tossed in the air and produced a “land” observatiion. Assume that each globe was equally likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on seeing “land” (Pr(Earth|land)), is 0.23.

```{r}

# Use the Bayes formula!

# probability of land, given Earth
p_le <- 0.3

# probability of land, given Mars
p_lm <- 1.0

# probability of Earth
p_e <- 0.5

# probability of land
p_l <- (p_e * p_le) + ((1 - p_e) * p_lm)

# probability of Earth, given land (using Bayes' Theorem)
p_el <- (p_le * p_e) / p_l
p_el



```

For the exercises below, I highly suggest you to grab a piece of paper and try to solve the problems in a 'visual' way.

2M4. Suppose you have a deck with only three cards. Each card has only two sides, and each side is either black or white. One card has two black sides. The second card has one black and one white side. The third card has two white sides. Now suppose all three cards are placed in a bag and shuffled. Someone reaches into the bag and pulls out a card and places it flat on a table. A black side is shown facing up, but you don’t know the color of the side facing down. Show that the probability that the other side is also black is 2/3. Use the counting method (Section 2 of the chapter) to approach this problem. This means counting up the ways that each card could produce the observed data (a black side faceing up on the table).

```{r}

# Write your probability here.

inner_tree <- 3/6
middle_tree <- 2/3

# Use the Bayes formula!

p_black_black <- 2/3

p_black_white <- 1/3

p_black <- 0.5

p_white <- (p_black * p_black_black) + ((1 - p_black) * p_black_white)

# probability of Earth, given land (using Bayes' Theorem)
p_black_black1 <- (p_black_black * p_black) / p_white
p_black_black1

#Shown w. right terms
card_bb_likelihood <- 2
card_bw_likelihood <- 1
card_ww_likelihood <- 0

likelihood <- c(card_bb_likelihood, card_bw_likelihood, card_ww_likelihood)
prior <- c(1, 1, 1)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)

posterior[1]

```

2M5. Now suppose there are four cards: B/B, B/W, W/W, and another B/B. Again suppose a card is drawn from the bag and a black side appears face up. Again calculate the probability that the other side is black.

```{r}

p_black_black <- 4/5

p_black_white <- 1/3

p_black <- 5/8

p_white <- (p_black * p_black_black) + ((1 - p_black) * p_black_white)

# probability of Earth, given land (using Bayes' Theorem)
p_black_black1 <- (p_black_black * p_black) / p_white
p_black_black1

#shown w. right terms
bb_likelihood <- 4
bw_likelihood <- 1
ww_likelihood <- 0

bb_prior <- 1
bw_prior <- 1
ww_prior <- 1

posterior_bb <- (bb_likelihood*bb_prior)/(bb_likelihood*bb_prior+bw_likelihood*bw_prior+ww_likelihood*ww_prior)
posterior_bb
```

2M6. Imagine that black ink is heavy, and so cards with black sides are heavier than cards with white sides. As a result, it’s less likely that a card with black sides is pulled from the bag. So again assume there are three cards: B/B, B/W, and W/W. After experimenting a number of times, you conclude that for every way to pull the B/B card from the bag, there are 2 ways to pull the B/W card and 3 ways to pull the W/W card. Again suppose that a card is pulled and a black side appears face up. Show that the probability the other side is black is now 0.5. Use the counting method, as before.

```{r}
bb_likelihood <- 2/3
bw_likelihood <- 1/3
ww_likelihood <- 0

bb_prior <- 1/6
bw_prior <- 2/6
ww_prior <- 3/6

bb_posterior <- (bb_likelihood*bb_prior)/(bb_likelihood*bb_prior+bw_likelihood*bw_prior+ww_likelihood*ww_prior)
bb_posterior

```
2M7. Assume again the original card problem, with a single card showing a black side face up. Before looking at the other side, we draw another card from the bag and lay it face up on the table. The face that is shown on the new card is white. Show that the probability that the first card, the one showing a black side, has black on its other side is now 0.75. Use the counting method, if you can. Hint: Treat this like the sequence of globe tosses, countng all the ways to see each observation, for each possible first card.

```{r}
# probability of black-black, given black
bb_b_likelihood <- 6/9

# probability of black-white, given black
bw_b_likelihood <- 3/9

# probability of black
b_prior <- 3/5 #due to one whit card is removed

# probability of white (marginalization)
p_w <- (bb_b_likelihood * b_prior) + ((1 - b_prior) * bw_b_likelihood)

# probability of Earth, given land (using Bayes' Theorem)
bb_b_posterior <- (bb_b_likelihood * b_prior) / p_w

bb_b_posterior
```
This task is best solved by drawing up the 'garden of forking paths' from the beginning of the chapter. Essentially, we count how many ways **each starting card** could have produced observation *W* on the second draw. 

Let's start with BB.

```{r, warning=FALSE}

library(dagitty)

# Create a dagitty object
garden <- dagitty("dag {
  BB -> B1
  BB -> B2
  B1 -> BW
  B1 -> WW
  B2 -> BW
  B2 -> WW
}")

# Plot it
plot(garden)

```

As we can see above, if our first card was BB, there are 6 ways the second draw could have been *W* - 3 for one side of BB (B1) and 3 for the other side of BB (B2).
 
Now let's assume the first card drawn was BW and count the ways for the second draw to be *W*.

```{r, warning=FALSE}

# Create our garden
garden <- dagitty("dag {
  BW -> B
  B -> WW
}")

# Plot it
plot(garden)

```

If the first card was BW, then the only way to get *W* on the second draw is by pulling the WW card out - either of it's side.

Now we table our findings.

| Cards | Ways to produce observation *W* on second draw |
|-------|------------------------------------------------|
| BB    | 6                                              |
| BW    | 2                                              |
| WW    | 0                                              |

Total ways to produce observation *W* on second draw: 8

Therefore, to get the probability that the first card drawn was BB is **6/8** or **0.75**.



    2H1. Suppose there are two species of panda bear. Both are equally common in the wild and live in the sample places. They look exactly alike and eat the same food, and there is yet no genetic assay capable of telling them apart. They differ however in their family sizes. Species A gives birth to twins 10% of the time, otherwise birthing a single infant. Species births twins 20% of the time, ottherwise birthing singleton infants. Assume these numbers are known with certainty, from many years of field research.
    Now suppose you are managing a captive panda breeding program. You have a newe female panda of unknown species, and she has just given birth to twins. What is the probability that her next birth will also be twins?

# After first birth, likelihood of species A and B is equal to the rate the
# species give birth to twins
a_likelihood <- 0.1
b_likelihood <- 0.2

# Next calculate the posterior probability that the panda belongs to each
# species, assume species are equally likely
likelihood <- c(a_likelihood, b_likelihood)
prior <- c(1, 1) 
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
posterior
#> [1] 0.333 0.667

# The probability the next birth is twins is the probability the panda belongs
# to each species times the likelihood each species gives birth to twins
(posterior[1] * a_likelihood) + (posterior[2] * b_likelihood)
#> [1] 0.167

    2H2. Recall all the facts from the problem above. Now compute the probability that the panda we have is from species A, asssuming we have observed only the first birth at that it was twins.

# probability of species A
p_a <- 0.5

# probability of twins, given species A
p_ta <- 0.1

# probability of twins, given species B
p_tb <- 0.2

# probability of twins
p_t <- (p_a * p_ta) + ((1 - p_a) * p_tb)

# probability of species A, given twins (using Bayes' Theorem)
# (note this is equivalent to `posterior[1]` above)
p_at <- (p_ta * p_a) / p_t
p_at
#> [1] 0.333

    2H3. Continuing on from the previous problem, suppose the same panda mother has a second birth and that it is not twins, but a singleton infant. Compute the posterior probability that this panda is species A.

# likelihood for each species is Pr(twins) * Pr(singleton)
a_likelihood <- 0.1 * (1 - 0.1)
b_likelihood <- 0.2 * (1 - 0.2)

# compute posterior probabilities
likelihood <- c(a_likelihood, b_likelihood)
prior <- c(1, 1)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)

posterior[1]
#> [1] 0.36

    2H4. A common boast of Bayesian statisticians is that Bayesian inference makes it easy to use all of the data, even if the data are of different types.
    So suppose now that a veterinarian comes along who has a new genetic test that she claims can identify the species of our mother panda. But the test, like all tests, is imperfect. This is the information you have about the test:

        The probability it correctly identifies a species A panda is 0.8.
        The probability it correctly identifies a species B panda is 0.65.

    The vet administers the test to your panda and tells you that the test is positive for species A. First ignore your previous information from the births and compute the posterior probability that your panda is species A. Then redo your calculation, now using the birth data as well.

# use Bayes' Theorem to determine the probability of species A, given a positive
# test
p_ap <- (0.8 * 0.5) / ((0.5 * 0.8) + (0.5 * 0.35))
p_ap
#> [1] 0.696


# Now include test data with observed births
# likelihood for each species is Pr(twins) * Pr(singleton)
a_likelihood <- 0.1 * (1 - 0.1)
b_likelihood <- 0.2 * (1 - 0.2)

# compute posterior probabilities, using test result as prior
likelihood <- c(a_likelihood, b_likelihood)
prior <- c(p_ap, (1 - p_ap))
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)

posterior[1]
#> [1] 0.563